---
title: "final_dw_distribution"
author: "abbey yatsko"
date: "7/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# SERC deadwood project
### script topic: deadwood survey exploration
deadwood >10cm from the forestGEO plot at SERC was surveyed over three time points (2014, 2017, 2021). for each survey, collected information varied, including the metrics / measurement scale for decay class. for example, decay class was measured as a 3-tiered variable in 2014, and 5-tiered variable in 2017 and 2021. based on classification descriptions from each protocol, the 5-tiered scale can be translated and broken down into that of the 3-tiered scale. this will be necessary to do when making tri-year comparisons of decay stage.  

additional metrics measured include: DBH, log/snag, height/length, top diameters (for downed pieces), air/ground contact, position within subplots, A/B/C pieces (for fragments) and so on... some of these variables also changed across years. of greatest interest are the dimensional measurements, decay class, species, and then differentiation between snags and logs.

following the 2021 survey, data comes from four sources: 
  * dead_2019.csv
  * living_2019.csv
  * deadwood_2014.csv
  * deadwood_2017.csv 

the goal of the study is to track pieces that cover all three time points of the survey and then sample these pieces for microbes/ C gas fluxes. the goal of this script is to parse through how dw distributions looked at each timepoint, and also how individual pieces move through 'the system' in time 

#### 1. ORGANIZE WORKSPACE----
```{r workspace}
library(ggplot2)
library(ggpubr)
library(dplyr)
library(tidyverse)

setwd("C:/Users/ary24/Desktop/repos/serc_deadwood/FINAL_DATA") # for lab server workflow
# setwd("/Users/abbeyyatsko/Desktop/repos/serc_deadwood/FINAL_DATA") #for macbook workflow 

dw2014 <- read.csv("deadwood_2014.csv", na.strings=c("","NA"))
dw2017 <- read.csv("deadwood_2017.csv", na.strings=c("","NA"))
dead_2019 <- read.csv("dead_2019.csv", na.strings=c("","NA"))
living_2019 <- read.csv("living_2019.csv", na.strings=c("","NA"))
```

note the initial length of the dataframes
  * 2014 has 1371 observations
  * 2017 has 682 observations (approximately half of the preceding survey)
  * dead_2019 has 460 observations
  * living_2019 has 4662 observations (most of these are going to be living trees, of which a handful had died and then moved into the           deadwood pool during the 2021 survey)

#### 2. EXPLORE DATA: dw2014 ----
```{r explore data dw2014}
dw2014 %>% summarise_all(n_distinct)
# 6 columns have only 1 unique value - these are filled with NAs. therefore, delete these rows all together, as they contain no information for us (just to simplify the dataset)
# therefore, get rid of columns that are completely empty
all_na <- function(x) any(!is.na(x))
dw2014 <- dw2014 %>% select_if(all_na)
# 6 removed columns specifically include: STEMTAG, SPOCDE, QX, QY, HEIGHT.2021, ANGLE
# dimensions are now 1371 observations of 19 variables 

# next step is to change TAG --> STEMTAG (to keep them the same across all 3 surveys, as this will be the link across all data frames for the merge)
dw2014 <- rename(dw2014, STEMTAG=TAG)

# rename Breakdownstatus.2014 --> BDS.2014
dw2014 <- rename(dw2014, BDS.2014=BREAKDOWNSTATUS.2014)

# rename decay class column to indicate year 2014
dw2014 <- rename(dw2014, DC.2014=DECAYCLASSORIG)

# rename decay class column to indicate year 2021 and include new abbreviation  
dw2014 <- rename(dw2014, DC.2021=DECAYCLASS.2021)

# rename DBH to indicate year 2014
dw2014 <- rename(dw2014, DBH.2014=DBH)

# drop direction column - only one val recorded and I think this was right at the beginning of the 2021 resurvey (we did not end up recording this across the 2021 survey)
dw2014 <- select(dw2014, -DIRECTION)
```

#### 3. EXPLORE DATA: dw2017 ----
```{r explore data dw2017}
dw2017 %>% summarise_all(n_distinct)
# looks like the X, X.1, X.2, X.3 variables are just empty with NAs. therefore, choose to remove these blank data columns - should reduce dimensions to 40 variables 
dw2017 <- dw2017 %>% select_if(all_na)
# good 

# rename Breakdownstatus.2017 --> BDS.2017
dw2017 <- rename(dw2017, BDS.2017=BREAKDOWNSTATUS.2017)

# rename decay class column to indicate year 2017 and include new abbreviation  
dw2017 <- rename(dw2017, DC.2017=DECAYCLASS.2017)

# get rid of DECAYCLASSORIG since it is the 'translated' version of DC.2017 from a 5-tier to a 3-tier decay classification system. this will be onboarded in later steps 
dw2017 <- select(dw2017, -DECAYCLASSORIG)

# rename decay class column to indicate year 2021 and include new abbreviation  
dw2017 <- rename(dw2017, DC.2021=DECAYCLASS.2021)

# QUESTION - what is the DBH column referring to? I don't think it matches up to 2014... leaving as is for now. TASK - assign meaning to this variable within the context of all surveys 

# drop direction and angle column - only one val recorded (ANGLE) and 9 in DIRECTION, we recorded this a few times at the beginning of the 2021 survey but then dropped it because the information was not deemed useful for my bigger picture decay class questions
dw2017 <- select(dw2017, -DIRECTION, -ANGLE)

# last few columns - I don't really know what these ones mean
# listed: LG.END.DIAM SM.END.DIAM LG.PX LG.PY SM.PX SM.PY CORRXTREE CORRYTREE UTMX_TREES UTMY_TREES
# keep for now, but note clutter
```

#### 4. EXPLORE DATA: dead_2019 ----
essentially what we have in this dataframe is a bunch of individuals that were dead in 2019. in the dw survey for 2021, we did not pick every last one of these up (visual inspection of spreadsheet). however, when new pieces were added to this, we recorded BDS, decay class, and dimensional measurements. all other observations may be coming from 2017 and 2014 surveys. 

```{r explore data dead_2019}
dead_2019 %>% summarise_all(n_distinct)
# only one column (ANGLE) without any content (only NAs). therefore, this will be pulled out. 
dead_2019 <- dead_2019 %>% select_if(all_na)
# good 

# QUESTION - what is the DBH column referring to? similar question to what shows up in dw2017... leaving as is for now. TASK - assign meaning to this variable within the context of all surveys 

# rename integrity.2021 --> integ (to keep consistent across data sets)
dead_2019 <- rename(dead_2019, INTEG.2021=INTEGRITY.2021)

# rename decay class column to indicate year 2021 and include new abbreviation  
dead_2019 <- rename(dead_2019, DC.2021=DECAYCLASS.2021)

# rename length column to indicate year 2021 
dead_2019 <- rename(dead_2019, LENGTH.2021=LENGTH)

# cut out direction - we recorded this a few times at the beginning of the 2021 survey but then dropped it because the information was not deemed useful for my bigger picture decay class questions
dead_2019 <- select(dead_2019, -DIRECTION)
```

#### 5. EXPLORE DATA: living_2019 ----
for this one, the majority of observations are actually for live things. therefore, we only want to pull out what we added in the 2021 dw survey, i.e. things that had died since the demography survey in 2019. these entries are indicated via those with BDS, decay class, any .2021 data essentially. let's filter to begin with by pulling out only stemtags that have a recorded value for BDS.2021 

```{r explore data living_2019}
# filter out living stems (those in which no info was recorded for in the 2021 dw survey). select for those samples that have recorded values for BDS.2021
living_2019 <- living_2019 %>%                                       
  filter(!is.na(BDS.2021)) 
# now what we have is all of the new recruits of dw since the 2019 survey 

living_2019 %>% summarise_all(n_distinct)
# looks like it makes sense to drop angle and direction again - do this in a few lines to keep things in order 

# rename integrity.2021 --> integ (to keep consistent across data sets)
living_2019 <- rename(living_2019, INTEG.2021=INTEGRITY.2021)

# rename decay class column to indicate year 2021 and include new abbreviation  
living_2019 <- rename(living_2019, DC.2021=DECAYCLASS.2021)

# rename length column to indicate year 2021 
living_2019 <- rename(living_2019, LENGTH.2021=LENGTH)

# drop direction and angle column - all NAs
living_2019 <- select(living_2019, -DIRECTION, -ANGLE)
```

now we should have all our data in a more meaningful, managable, and consistent form (when it comes to variables especially). here they are:

```{r list polished data frames}
View(dw2014)
View(dw2017)
View(dead_2019)
View(living_2019)
```

QUESTION: 
from dw_2014 should we remove rows that are recorded with 'no tag' for the stemtag variable? having no stem tag is not useful for tracking individual pieces thru time, but could be helpful for overall biomass estimates (especially if some of these no-taggers are big trees that died prior to forestGEO censusing - representing large C sinks thru biomass, even if untracked/unidentifiable at the species level)

#### 6. MERGING DATA ----
the next goal would be to merge all of the dataframes across the variable 'STEMTAG'. first we need to check out how the variables are being understood (make sure they are the same across all 4 sets)

```{r variable classes, results='hide'}
str(dw2014)
dw2014$QUADNAME <- as.factor(dw2014$QUADNAME)
dw2014$STEMTAG <- as.factor(dw2014$STEMTAG)
dw2014$DC.2014 <- as.factor(dw2014$DC.2014)
dw2014$DC.2021 <- as.factor(dw2014$DC.2021)
dw2014$TOPDIAM.2021 <- as.numeric(dw2014$TOPDIAM.2021)
dw2014$LENGTH.2014 <- as.numeric(dw2014$LENGTH.2014)

str(dw2017)
dw2017$QUADNAME <- as.factor(dw2017$QUADNAME)
dw2017$TAG <- as.factor(dw2017$TAG)
dw2017$STEMTAG <- as.factor(dw2017$STEMTAG)
dw2017$DC.2017 <- as.factor(dw2017$DC.2017)
dw2017$DC.2021 <- as.factor(dw2017$DC.2021)

str(dead_2019)
dead_2019$QUADNAME <- as.factor(dead_2019$QUADNAME)
dead_2019$TAG <- as.factor(dead_2019$TAG)
dead_2019$STEMTAG <- as.factor(dead_2019$STEMTAG)
dead_2019$INTEG.2021 <- as.character(dead_2019$INTEG.2021)
dead_2019$DC.2021 <- as.factor(dead_2019$DC.2021)

str(living_2019)
living_2019$QUADNAME <- as.factor(living_2019$QUADNAME)
living_2019$TAG <- as.factor(living_2019$TAG)
living_2019$STEMTAG <- as.factor(living_2019$STEMTAG)
living_2019$DC.2021 <- as.factor(living_2019$DC.2021)
```

before merging, see what the common elements are across: 
  * 2014 -> 2017 -> 2021
  * 2014 -> 2017
  * 2014 -> 2021
  * 2017 -> 2021
  
```{r common elements}
# > dw2014
dw2014_tagvec <- dplyr::pull(dw2014, STEMTAG)
# > dw2017
dw2017_tagvec <- dplyr::pull(dw2017, STEMTAG)
# > dead_2019
dead_2019_tagvec <- dplyr::pull(dead_2019, STEMTAG)
# > living_2019 
living_2019_tagvec <- dplyr::pull(living_2019, STEMTAG)

# common elements 2014 -> 2017
ce_14_17 <- Reduce(intersect, list(dw2014_tagvec, dw2017_tagvec))
ce_14_17
# there are 169 stemtags that are represented across 2014 and 2017. that's cool. 

# what species are these shared stemtags (2014-2017)

# QUESTION: is it more useful to have a search that pulls out unique stem tags that have records for DC.2014, DC.2017, and DC.2021? this is more so what we are interested in... especially since the 4 seperate dataframes do not actually translate to things from the 2014, 2017, 2021 survey specifically.

```
prep data to merge - need to make sure that all of the columns that need to be 'shared' are included individually in each of the 4 merging datasets. define these as mergeCols: 
 [1] "QUADNAME"     "STEMTAG"      "BDS.2021"     "PIECE.2021"   "INTEG.2021"   "DC.2021"     
 [7] "GC.2021"      "TOPDIAM.2021" "DBH.2021"     "LENGTH.2021" 

I think we also want to add HEIGHT.2021 - it is shared across all 3 minus the dw2014. just add the column 
```{r pre-merge}
# add HEIGHT.2021 to dw2014 and populate with NAs - just so that we can get this variable across all 4 dataframes 
# also do this for SPCODE

dw2014$HEIGHT.2021 <- 'NA'
str(dw2014)
dw2014$HEIGHT.2021 <-  as.numeric(dw2014$HEIGHT.2021)
dw2014$SPCODE <- NA
dw2014$SPCODE <-  as.character(dw2014$SPCODE)

# create the merging parameters 
list_df = list(dw2014,dw2017,dead_2019,living_2019)

col_common = colnames(list_df[[1]])
for (i in 2:length(list_df)){
  col_common = intersect(col_common, colnames(list_df[[i]]))
}
col_common


```

ok now for actual merge based on the col_common variable 
  * note that
```{r merge}
# for dw2014 and dw2017 
dw2014 %>%
  full_join(dw2017, by = col_common)%>%
  full_join(dead_2019, by = col_common)%>%
  full_join(living_2019, by = col_common) -> dw_merged
View(dw_merged)
str(dw_merged)
```

#### 7. STARTING WITH VISUALIZATIONS ----
ok so if I did this right (let's pretend), we could theoretically start making sense of the data and doing some viz with it?? 

```{r initial viz}
# what kind of species are we working with? 
dw_merged %>% 
        drop_na(SPCODE) %>%
        ggplot(aes(x = SPCODE))+
        geom_bar(stat="count")+
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# lots of ash, sweetgum, tulip tree... think about what species to target here?

# what is the distribution of decay classes across 2014, 2017, and 2021? while also taking into account the different breakdown status... 

dc2014 <- ggplot(data = subset(dw_merged, !is.na(DC.2014)), aes(fill=BDS.2014, x=DC.2014)) + 
    geom_bar(position="stack", stat="count")

dc2017 <- ggplot(data = subset(dw_merged, !is.na(DC.2017)), aes(fill=BDS.2017, x=DC.2017)) + 
    geom_bar(position="stack", stat="count")

dc2021 <- ggplot(data = subset(dw_merged, !is.na(DC.2021)), aes(fill=BDS.2021, x=DC.2021)) + 
    geom_bar(position="stack", stat="count")

ggarrange(dc2014, dc2017, dc2021, nrow = 1)

# also look at DC distribution but now take into account species...  but just for 2017 and 2021 becuase theres not actually species data that is paired with the 2014 pieces 

dc2017_sp <- ggplot(data = subset(dw_merged, !is.na(DC.2017)), aes(fill=SPCODE, x=DC.2017)) + 
    geom_bar(position="stack", stat="count")

dc2021_sp <- ggplot(data = subset(dw_merged, !is.na(DC.2021)), aes(fill=SPCODE, x=DC.2021)) + 
    geom_bar(position="stack", stat="count")

ggarrange(dc2017_sp, dc2021_sp, nrow = 1)
```

#### 8. NEXT STEPS AND THINGS I WANT TO KNOW ----
  * need to pull out samples that account for the following transitions: 
    * 2014 -> 2017
    * 2014 -> 2017 -> 2021
    * 2014 -> 2021
  * thinking about transition matrix and how to actually code this 
  * make sure that the dw_merged dataframe is what I think it is 
  * best way to make rules for using dimensional measurements to construct biomass estimates? 
  
##### **THE BIG GOAL IS TO FIGURE OUT WHAT PIECES WE WANT TO TARGET FOR SUBSAMPLING C FLUX AND MICROBES!**

pull out only sample IDs and then decay classes across the three years 
```{r decay class dataframe}
dw_merged_dc <- dw_merged %>%
  select(c('STEMTAG', 'DC.2014', 'DC.2017', 'DC.2021'))


```


