---
title: "final_dw_distribution"
author: "abbey yatsko"
date: "7/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# SERC deadwood project
### script topic: deadwood survey exploration
deadwood >10cm from the forestGEO plot at SERC was surveyed over three time points (2014, 2017, 2021). for each survey, collected information varied, including the metrics for decay class. decay class was measured as a 3-tiered variable in 2014, and 5-tiered variable in 2017 and 2021. based on classification descriptions from each protocol, the 5-tiered scale can be translated and broken down into that of the 3-tiered scale 

additional metrics measured include: DBH, log/snag, height/length, top diameters (for downed pieces), air/ground contact, position within subplots, A/B/C pieces (for fragments) and so on... 

following the 2021 survey, data comes from four sources: 
  * dead_2019.csv
  * living_2019.csv
  * deadwood_2014.csv
  * deadwood_2017.csv 

the goal of the study is to track pieces that cover all three time points of the survey and then sample these pieces for microbes/fluxes. the goal of this script is to parse through how dw distributions looked at each timepoint, and also how individual pieces move through 'the system' in time 

#### 1. ORGANIZE WORKSPACE----
```{r workspace}
library(ggplot2)
library(ggpubr)
library(dplyr)
library(tidyverse)

setwd("/Users/abbeyyatsko/Desktop/repos/serc_deadwood/FINAL_DATA")
dw2014 <- read.csv("deadwood_2014.csv", na.strings=c("","NA"))
dw2017 <- read.csv("deadwood_2017.csv", na.strings=c("","NA"))
dead_2019 <- read.csv("dead_2019.csv", na.strings=c("","NA"))
living_2019 <- read.csv("living_2019.csv", na.strings=c("","NA"))
```

note the initial length of the dataframes
  * 2014 has 1371 observations
  * 2017 has 682 observations (approximately half of the preceding survey)
  * dead_2019 has 460 observations
  * living_2019 has 4662 observations

#### 2. EXPLORE DATA: dw2014 ----
```{r explore data dw2014}
dw2014 %>% summarise_all(n_distinct)
# 6 columns have only 1 unique value - these are filled with NAs. therefore, delete these rows all together, as they contain no information for us (just to simplify the dataset)
# therefore, get rid of columns that are completely empty
all_na <- function(x) any(!is.na(x))
dw2014 <- dw2014 %>% select_if(all_na)
# 6 removed columns specifically include: STEMTAG, SPOCDE, QX, QY, GC.2021, ANGLE
# dimensions are now 1371 observations of 19 variables 

# next step is to change TAG --> STEMTAG (to keep them the same across all 3 surveys, as this will be the link across all data frames for the merge)
dw2014 <- rename(dw2014, STEMTAG=TAG)

# rename Breakdownstatus.2014 --> BDS.2014
dw2014 <- rename(dw2014, BDS.2014=BREAKDOWNSTATUS.2014)

# rename decay class column to indicate year 2014
dw2014 <- rename(dw2014, DC.2014=DECAYCLASSORIG)

# rename decay class column to indicate year 2021 and include new abbreviation  
dw2014 <- rename(dw2014, DC.2021=DECAYCLASS.2021)

# rename DBH to indicate year 2014
dw2014 <- rename(dw2014, DBH.2014=DBH)

# drop direction column - only one val recorded and I think this was right at the beginning of the 2021 resurvey (we did not end up recording this across the 2021 survey)
dw2014 <- select(dw2014, -DIRECTION)
```










